A framework for linking dynamic experiences to memories reveals event-like structure in naturalistic episodic recall

### 2019-06-19 paper/main.tex
finished edits to text, a few references left

%\title{How is experience transformed into memory?}
%\title{A content-based framework for capturing stimulus and memory dynamics reveals event-like structure in naturalistic episodic recall}
%\title{A content-based model characterizing/describing/representing how experiences and memories unfold over time reveals event-like structure in naturalistic episodic recall}
%\title{A content-based framework describing the temporal dynamics of experiences and memories reveals event-like structure in naturalistic episodic recall}
\title{A content-based model of how experiences and memories unfold over time reveals event-like structure in naturalistic episodic recall}

@@

\begin{abstract}
The ways our experiences unfold over time define unique \textit{trajectories} through the relevant representational spaces.  Within this geometric framework, one can compare the shape of the trajectory formed by an experience to that defined by our later remembering of that experience.  We propose a framework for mapping naturalistic experiences onto geometric spaces that characterize how they unfold over time.  We apply this approach to a naturalistic memory experiment which had participants view and recount a video.  We found that the shapes of the trajectories formed by participants' recountings were all highly similar to that of the original video, but participants differed in the level of detail they remembered.  Further, we identified a network of brain structures that are sensitive to the ``shapes'' of our ongoing experiences, and an overlapping network that is sensitive to how we will later remember those experiences.
\end{abstract}

@@

What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from the rest of one's experiences, and that item may be labeled as having been recalled versus forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. Using well-controlled, trial-based experimental designs, the field has amassed a wealth of valuable information regarding human episodic memory.  However, there are fundamental properties of our memories that trial-based experiments are not well suited to capture~\citep[for review also see][]{KoriGold94, HukEtal18}.  First, our memories are continuous, rather than discrete: removing a (naturalistic) event from the context in which it occurs can substantially change its meaning.  Second, the specific words used to describe an experience have little bearing on whether the experience should be considered to have been ``remembered.''  Asking whether the rememberer has precisely reproduced a specific set of words to describe a given experience is nearly orthogonal to whether they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, counting the number or proportion of precise recalls is often a primary metric of assessing the quality of participants' memories.  Third, one might remember the essence of an experience but forget (or neglect to recount) particular details.  Capturing the essence (or the \textit{shape}) of what happened is typically the main ``point'' of recounting a memory to a listener.

How might one go about formally characterizing the shape of an experience, or whether that shape has been recovered by the rememberer?  Any given moment of an experience derives meaning from surrounding moments, as well as from longer-range temporal associations~\citep[e.g., ][]{LernEtal11}.  Therefore the timecourse describing how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g., ][]{HowaKaha02, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15}.  Our memory systems can then leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g., the room or situation we are in is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.  Although our experiences most often change gradually, they also occasionally change suddenly~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) during an experience help to discretize our experiences into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within event) and transient (across event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distill those experiences down to their essence.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}. Prior studies have implicated the hippocampus and the medial prefrontal cortex as playing a critical role in this process [CITE].

Here we sought to examine how the temporal dynamics of a ``naturalistic'' experience were reflected in participants' later memories of that experience.  We analyzed an open dataset that comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recalled an episode of the BBC television series \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode and of participants' verbal recalls.  Specifically, we use topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls, and we use Hidden Markov Models~\citep{Rabi89, BaldEtal17} to discretize the evolving semantic content into events.  In this way, we cast naturalistic experiences (and recalls of those experiences) as \textit{topic trajectories} that describe how the experiences evolve over time.  In other words, the episode's topic trajectory is a formalization of its shape.  Under this framework, successful remembering entails verbally ``traversing'' the topic trajectory of the original episode, thereby reproducing the original episode's shape.  In addition, comparing the shapes of the topic trajectories of the original episode and of participants' retellings of the episode reveals which aspects of the episode were preserved (or lost) in the translation into memory.  Further, we identified a network of brain structures whose responses (as participants watched the episode) reflected the shape of the episode, and a second network whose responses reflected how participants would later recount the episode.

@@

Statistical models of memory studies often treat memory recalls as binary (e.g. the item was recalled or not) and independent events.  However, our approach produces a content-based model of individual stimulus events and recall events, allowing for direct quantitative comparison between all stimulus and recall events, as well as between the recall events themselves.  Leveraging this affordance, we introduce 2 novel metrics for quantifying memory representations: precision and distinctiveness.  We define precision as the average correlation between each recall event and the maximally correlated video event (Fig.~\ref{fig:precision-distinctiveness}).  Participants whose recall events are more veridical descriptions of what happened in the video event will presumably have higher precision scores. We find that across participants, a higher precision score is correlated to both hand annotated memory performance (Pearson's $r(15) = 0.6, p = 0.011$) as well as the number of recall events (per participant) estimated by our model (Pearson's $r(15) = 0.64, p = 0.005$). A second novel metric we introduce here is the distinctiveness, or the uniqueness of each recall event relative to other recall events. Distinctiveness is defined as 1 minus the average of the upper triangle of the recall-by-recall correlation matrix. Similar to precision, we find that the more distinct participants recalls are (on average), the more they remembered (hand-annotated memory: Pearson's $r(15) = 0.83, p < 0.001$ and model derived memory: Pearson's $r(15) = 0.71, p = 0.001$).  In summary, using two novel metrics afforded by our approach, we find that participants whose recalls are both more precise and distinct remember more content.

@@

Many statistical models of memory treat recall as binary and unrelated events. Our novel framework allows one to measure the \textit{precision} of each recall event (by considering the correlation to the highest video event) as well as how \textit{distinctive} each recall event is from other recall events.  These new metrics afforded by our method allow for much more nuanced analysis of recall dynamics than were previously possible.  Using them, we find that both precision and distinctiveness are correlated to the amount of information participants recall. Thus, the manner in which a person recounts a \textit{particular} event is related to their ability to recall \textit{many} events.

While a large number of language models exist (e.g. WAS, LSA, word2vec, universal sentence encoder), here we use topic models for a few reasons.  First, topic models capture the \textit{essence} of a text passage devoid of the specific set and order of words used.  This was an important feature of our model since different people may accurately recall a scene using very different language. Secondly, words can mean different things in different contexts (e.g. baseball bat vs. the animal bat), and topic models are robust to this since words can be part of multiple topics.  Lastly, topic models provide a straight forward to recover the weights for the particular words comprising a topic, allowing for easy interpretation of an event's contents (e.g. Fig.~\ref{fig:topics}). Other models such as Google's universal sentence encoder offer a context-sensitive encoding of text passages, but the encoding space is complex and non-linear and thus, recovering the original words used to fit the model is not straight forward.

@@

Statistical models of memory studies often treat memory recalls as binary and distinct events.  However, our approach produces a formal model of individual stimulus events and recall events, allowing for direct quantitative comparison between all stimulus and recall events, as well as between the recall events themselves.  Leveraging this fact, we introduce 2 novel metrics for quantifying memory representations: precision and distinctiveness.

\paragraph*{Precision.}
We next tested whether participants who recalled more events were also more \textit{precise} in their recollections. For each participant, we computed the correlation between the topic vectors for each recall event and that of its closest-matching video event (only for the events which they recalled). We defined the precision as the average video-recall correlation across all of the events a participant recalled.  We found a strong correlation between hand-annotated memory performance and precision, suggesting that participants who remembered more events also remembered them more veridically (Pearson's $r(15) = 0.74, p = 0.0006$).

\paragraph*{Distinctiveness.}
We also considered the \textit{distinctiveness} of each recalled event. That is, how uniquely a recalled event's topic vector matched a given video event topic vector, versus the topic vectors for the other video events. We hypothesized that participants with high memory performance might describe each event in a more distinctive way (relative to those with lower memory performance who might describe events in a more general way).  To test this hypothesis we define a distinctiveness score for each recalled event as
\[
  d(\mathrm{event}) = 1 - \bar{c}(\mathrm{event}),
\]
where $\bar{c}(\mathrm{event})$ is the average correlation between the given recalled event's topic vector and the topic vectors from all video events \textit{except} the best-matching video event.  We then averaged these distinctiveness scores across all of the events recalled by the given participant.  We found that participants with higher average distinctiveness scores tended to also have better hand-annotated memory performance (Pearson's $r(15) = 0.8, p = 0.0001$).

>>>>

%\title{How is experience transformed into memory?}
%\title{A content-based framework for capturing stimulus and memory dynamics reveals event-like structure in naturalistic episodic recall}
%\title{A content-based model characterizing/describing/representing how experiences and memories unfold over time reveals event-like structure in naturalistic episodic recall}
%\title{A content-based framework describing the temporal dynamics of experiences and memories reveals event-like structure in naturalistic episodic recall}
%\title{A content-based model of how experiences and memories unfold over time reveals event-like structure in naturalistic episodic recall}
\title{A framework for linking dynamic experiences to memories reveals event-like structure in naturalistic episodic recall}

@@

\begin{abstract}
Our life experiences unfold over time and the temporal dynamics of their contents form unique \textit{experience trajectories}.  Within this geometric framework, one can compare the shape of the trajectory formed by an experience to that defined by our later remembering of that experience.  We propose a framework for mapping naturalistic experiences onto geometric spaces that characterize how they unfold over time.  We apply this approach to a naturalistic memory experiment which had participants view and recount a video.  We find that the video and subsequent recall share an event-like structure, and that the shapes of the trajectories formed by participants' recountings were all highly similar to that of the original video. Further, the level of precision that participants' recount the events as well as how distinct their recountings are from other recall events both predict overall subsequent memory performance. Lastly, we identified a network of brain structures that are sensitive to the ``shapes'' of our ongoing experiences, and an overlapping network that is sensitive to how we will later remember those experiences. These result highlight that our dynamic experiences are structured into events in memory, and introduce a novel framework for mapping between dynamic life experiences and their mnemonic counterparts.
\end{abstract}

@@

What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from the rest of one's experiences, and that item may be labeled as having been recalled versus forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. Undoubtedly, using well-controlled trial-based experimental designs, the field has amassed a wealth of valuable information regarding human episodic memory.  However, there are fundamental properties of our memories that trial-based experiments are not well suited to capture~\citep[for review also see][]{KoriGold94, HukEtal18}.  First, our memories are continuous, rather than discrete: removing a (naturalistic) event from the context in which it occurs can substantially change its meaning.  Second, the specific language used to describe an experience have little bearing on whether the experience should be considered to have been ``remembered.''  Asking whether the rememberer has precisely reproduced a specific set of words to describe a given experience is nearly orthogonal to whether they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, counting the number or proportion of precise recalls is often a primary metric of assessing the quality of participants' memories.  Third, one might remember the \textit{essence} (or shape) of an experience but forget (or neglect to recount) particular details.  Capturing the essence of what happened is typically the main ``point'' of recounting a memory to a listener.

How might one go about formally characterizing the shape of an experience, or whether that shape has been recovered by the rememberer?  Any given moment of an experience derives meaning from surrounding moments, as well as from longer-range temporal associations~\citep[e.g., ][]{LernEtal11}.  Therefore, the timecourse describing how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g., ][]{HowaKaha02, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15}.  Our memory systems can then leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g., the room or situation we are in is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.

Although our experiences most often change gradually, they also occasionally change suddenly~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) during an experience help to discretize our experiences into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within event) and transient (across event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distill those experiences down to their essence.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}. Prior research has implicated the hippocampus and the medial prefrontal cortex as playing a critical role in transforming experiences into stuctured and consolidated memories ~\citep{TompDava17}.

Here we sought to examine how the temporal dynamics of a ``naturalistic'' experience were reflected in participants' later memories of that experience.  We analyzed an open dataset that comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recalled an episode of the BBC television series \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode and of participants' verbal recalls.  Specifically, we use topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls, and we use Hidden Markov Models~\citep{Rabi89, BaldEtal17} to discretize the evolving semantic content into events.  In this way, we cast naturalistic experiences (and recalls of those experiences) as \textit{trajectories} that describe how the experiences evolve over time. Under this framework, successful remembering entails verbally ``traversing'' the topic trajectory of the original episode, thereby reproducing the shape of the original experience.  Comparing the shapes of the topic trajectories of the original episode and of participants' retellings of the episode reveals which aspects of the episode were preserved (or lost) in the translation into memory.

% NOTE: Can we remove this part? its captured in the abstract
% Using a series of novel memory metrics (introduced here and afforded by our computational framework), we find that the successful memory performance is related to 1) the \textit{precision} with which the participant recounts each event and 2) how \textit{distinctive} each recall event is (relative to the other recalled events). Finally, we identified a network of brain structures whose responses (as participants watched the episode) reflected the shape of the episode, and a second network whose responses reflected how participants would later recount the episode.

@@

Statistical models of memory studies often treat memory recalls as binary (e.g. the item was recalled or not) and independent events.  However, our framework produces a content-based model of individual stimulus events and recall events, allowing for direct quantitative comparison between all stimulus and recall events, as well as between the recall events themselves.  Leveraging this affordance, we introduce 2 novel metrics for quantifying naturalistic memory representations: precision and distinctiveness.  We define precision as the average correlation between each recall event and the maximally correlated video event (Fig.~\ref{fig:precision-distinctiveness}).  Participants whose recall events are more veridical descriptions of what happened in the video event will presumably have higher precision scores. We find that across participants, a higher precision score is correlated to both hand annotated memory performance (Pearson's $r(15) = 0.6, p = 0.011$) as well as the number of recall events (per participant) estimated by our model (Pearson's $r(15) = 0.64, p = 0.005$). A second novel metric we introduce here is the distinctiveness, or the uniqueness of each recall event relative to other recall events. We define distinctiveness as 1 minus the average of all non-matching recall events from the video-recall correlation matrix. We hypothesized that participants who recounted events in a more distinctive way would display better overall memory.  Similar to precision, we find that the more distinct participants recalls are (on average), the more they remembered (hand-annotated memory: Pearson's $r(15) = 0.83, p < 0.001$ and model derived memory: Pearson's $r(15) = 0.71, p = 0.001$).  In summary, using two novel metrics afforded by our approach, we find that participants whose recalls are both more precise and distinct remember more content.

@@

When modeling memory experiments, often times events (or items) and their later memories are treated as binary (or categorical in the case of confidence ratings) and independent events. Our novel framework allows one to assess memory performance in a more continuous way, as well as analyze the correlational structure of each encoding event to each memory event. Further and importantly, it allows for consideration of the actual content of the experience/memories, which is not typically modeled. Leveraging this, using 2 novel memory metrics we find that the successful memory performance is related to 1) the \textit{precision} with which the participant recounts each event and 2) how \textit{distinctive} each recall event is (relative to the other recalled events). The first finding suggests to us that the accuracy of recall for \textit{any individual event} may predict the overall amount of information recovered by the participant.  The second finding suggests that remembering/describing events in a unique way (relative to other recalled events) is also related to the quantity of content recovered. Intriguingly, prior studies show that pattern separation, or the ability to discriminate between similar experiences, is impaired in many cognitive disorders as well as natural aging [Craig Stark and friends].  Future work might explore how/whether these metrics compare between cognitively impoverished groups and healthy controls.

While a large number of language models exist (e.g. WAS, LSA, word2vec, universal sentence encoder), here we use topic models for a few reasons [CITE THESE].  First, topic models capture the \textit{essence} of a text passage devoid of the specific set and order of words used.  This was an important feature of our model since different people may accurately recall a scene using very different language. Secondly, words can mean different things in different contexts (e.g. baseball bat vs. the animal bat), and topic models are robust to this since words can be part of multiple topics.  Lastly, topic models provide a straight forward to recover the weights for the particular words comprising a topic, allowing for easy interpretation of an event's contents (e.g. Fig.~\ref{fig:topics}). Other models such as Google's universal sentence encoder offer a context-sensitive encoding of text passages, but the encoding space is complex and non-linear and thus, recovering the original words used to fit the model is not straight forward.

@@

\paragraph*{Precision.}
We tested whether participants who recalled more events were also more \textit{precise} in their recollections. For each participant, we computed the correlation between the topic vectors for each recall event and that of its closest-matching video event (only for the events which they recalled). We Fisher's z-transformed the correlations, computed the average and then inverse Fisher's z-transformed the resulting value.  This gave a single value per participant representing the average precision across all recalled events.  We then correlated this value with hand-annotated as well as model derived (e.g. $k$ or the number of events recovered by the HMM) memory performance.

\paragraph*{Distinctiveness.}
We also considered the \textit{distinctiveness} of each recalled event. That is, how uniquely a recalled event's topic vector matched a given video event topic vector, versus the topic vectors for the other video events. We hypothesized that participants with high memory performance might describe each event in a more distinctive way (relative to those with lower memory performance who might describe events in a more general way).  To test this hypothesis we define a distinctiveness score for each recalled event as

%NOTE: Need to check on this equation!
\[
  d(\mathrm{event}) = 1 - \bar{c}(\mathrm{event}),
\]
where $\bar{c}(\mathrm{event})$ is the average correlation between the given recalled event's topic vector and the topic vectors from all video events \textit{except} the best-matching video event.  We then averaged these distinctiveness scores across all of the events recalled by the given participant. As above, we used Fisher's z (transform and inverse-transform) before/after averaging correlation values. Finally, we correlated these values with hand-annotated and model derived memory performance scores across-subjects.
